---
title: "Stat 331 Final Project"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---
# Introduction

```{r setup}
library(tidyverse)
```
 **Data and Variable Description**:
We chose to explore the relationship between child mortality rate and average daily income by country. In order to find data for this, we selected two datasets from Gapminder, one for each of our quantitative variables. For our first variable, we are measuring child mortality by country as the number of 0-5 year olds dying per 1000 babies born, and for income we are measuring average daily household per capita income adjusted for inflation and price. Without controlling for inflation, we would see a near constant increase in income by year, but by controlling for income by expressing every income in 2017 US dollars we can better compare different countries.

**Hypothesis:**: 
Our hypothesis is that we will see an inverse relationship between each countries child mortality rate and average daily income, or that as income increases, child mortality decreases by country. We would expect childcare and medical care to be better and more readily available in countries as income increases, decreasing the child mortality rate. Also, we would expect countries with higher income to have a lower child mortality rate. While there are numerous other confounding variables that also effect child mortality rate, we will only focus on the relationship between average daily income and child mortality rate.

# Section 1.1: Data Cleaning
```{r download data}

child_mortality <- read_csv("https://raw.githubusercontent.com/drewclanton/STAT331_Project/main/child_mortality_0_5_year_olds_dying_per_1000_born.csv")
income <- read_csv('https://raw.githubusercontent.com/drewclanton/STAT331_Project/main/mincpcap_cppp.csv')
```

```{r}
sum(is.na(income))
sum(is.na(child_mortality))
```


# Section 1.2: Pivoting longer
# Section 1.3: Joining Datasets

