---
title: "Stat 331 Final Project"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
editor: source
execute: 
  error: true
  echo: false
  message: false
  warning: false
---
# Introduction

```{r setup}
#| include: false
library(tidyverse)
library(gganimate)
library(gifski)
library(knitr)
library(kableExtra)
library(gridExtra)
```
 **Data and Variable Descriptions**:
 
We chose to explore the relationship between child mortality rates and average daily income by country from the years 1800 to 2023. In order to find data for this, we selected two datasets from Gapminder, one for each of our quantitative variables. For our first variable, we are measuring child mortality by country as the number of 0-5 year olds dying per 1000 babies born, and for income we are measuring average daily household per capita income adjusted for inflation and price. Without controlling for inflation, we would see a near constant increase in income by year, but by controlling for income by expressing every income in 2017 US dollars we can better compare different countries since we are standardizing the value of the dollar across all data points. 

It is also important to consider the origins and logic behind the procurement of these datasets as Gapminder has extrapolated and made rough estimates of changes in both of these datasets that are not necessarily accurate. Within the child mortality rates dataset, Gapminder combined data from multiple sources ranging in methodology as some of the mortality rates are calculated based on regression analysis while others are simply projected estimates. This, coupled with the varying yearly ranges leads to a patchwork of datasets to create the dataset we are analyzing which could eliminate the relevancy of our analysis. The mean income per capita dataset faces a similar problem as the various countrys' income distributions are calculated via different methods based on the prevalency of data. If countries do not have a mean income indicator from the World Bank with data from household surveys, daily incomes per household are instead estimated based on the country's GDP per capita. These factors should be considered in assessing the relevancy, accuracy, and value of our analysis between the two variables in question.

**Hypothesis:** 

  Our hypothesis is that we will see an inverse relationship between each countries child mortality rate and average daily income, or that as income increases, child mortality decreases by country. We would expect childcare and medical care to be better and more readily available in countries as income increases, decreasing the child mortality rate. A greater daily household income also indicates a larger economy with better technology to treat emergencies leading to improved citizen welfare. While there are numerous other confounding variables that also effect child mortality rate, we will only focus on the relationship between average daily income and child mortality rate, and it is important to acknowledge that our analysis may prove association but not necessarily causation between the two varaibles. 

# Section 1.1: Data Cleaning and Pivoting
```{r download data}
#| include: false
child_mortality <- read_csv("https://raw.githubusercontent.com/drewclanton/STAT331_Project/main/child_mortality_0_5_year_olds_dying_per_1000_born.csv")
income <- read_csv('https://raw.githubusercontent.com/drewclanton/STAT331_Project/main/mincpcap_cppp.csv')
```

Our goal in data cleaning is looking to create clean and tidy data that we can easily analyze. First, we checked for any missing values in the data:
```{r}
sum(is.na(income))
sum(is.na(child_mortality))
```
We found that our daily income data set (income) has zero missing values, but our child mortality data set is missing 1800 entries. 

We investigated the nature of our missing values, learning that they are not random, but rather some countries do not have any values for any years. These countries cannot be included because we need at least some entries to perform analysis on, so we will ommit them from our analysis. 

```{r}
#data cleaning on child mortality data set
child_mortality <- child_mortality |>
  na.omit() |>
  pivot_longer(cols =`1800`:`2100`,
               names_to = "year",
               values_to = "child_mortality") |>
  filter(year<2024)

#data cleaning on income set
income <- income |>
  pivot_longer(cols =`1800`:`2100`,
               names_to = "year",
               values_to = "daily_income") |>
  filter(year<2024)

  
```

Our next step was to pivot our data to a long format where we will have a single column for our variable of interest, meaning that each observation will have as its variables: Country, Year, and Variable of Interest (income or child mortality). This format will make our data easier to join as well as making it easier to filter out our specified years. Given the extrapolation in our data set, we will limit our observations to the period of 1800-2023, removing all future speculative entries. After this step, we are left with two much longer data sets, as each country will now have 2023 observations, one for each year. 


# Section 1.3: Joining Datasets

```{r}
data <- child_mortality |>
  left_join(income, join_by(country==country,
                            year==year))
```

In combining our data sets, we chose to only include entries that were present in the child mortality dataset with no missing values. Since our income data set has no missing values, we will not have to worry about any NAs in our data. We ended up with 185 countries in our final data set, meaning that we only removed 12 countries from the set for the reason of having no data about child mortality. As these countries have zero entries, it would be impossible to test our hypothesized relationship so we require countries with enough data to interpret.

# Section 2.1: Data Visualization 
Our first data visualization is a scatterplot that explores the relationship of child mortality and daily income over time. Each point on the plot represents a country's child mortality rate and daily income for an individual year. We do not see an exactly linear relationship, rather a slightly parabolic curve as child mortality rate decreases as daily income increases. There are few real outliers in this data, as most points are packed very tightly together.
```{r}
data |>
  ggplot(mapping=aes(y=child_mortality,x=daily_income)) +
  geom_point(color='steelblue') +
  theme_classic() +
  labs(title = "Child Mortality vs. Daily Income",
       x = "Daily Income",
       y = "Child Mortality")
```
While the above graph shows us the relationship of daily income and child mortality over all years for all countries, we are curious to see how the relationship between daily income and child mortality has changed over time. Below, is an animated scatterplot that displays the relationship between child mortality and daily income by year. We see that as more time passes, child mortality rate drops, and daily income starts to move higher up. We see that daily income increases the most after 1960, while child mortality continues to get closer to 0.
```{r}
time<- data |>
  mutate(year = as.integer(as.character(year))) |>
  ggplot(mapping=aes(x=daily_income,y=child_mortality)) + 
  geom_point(color='steelblue') +
  labs(title = 'Child Mortality and Daily Income by Country Animated by Year: {frame_time}',
       x = 'Daily Income',
       y = 'Child mortality',
       ) +
  transition_time(year) +
  ease_aes('linear') 
animate(time,renderer=gifski_renderer())
```

# Section 2.2: Linear Regression:

```{r}
mod_data <- data |>
  group_by(country) |>
  summarise(avg_mortality=mean(child_mortality),avg_income=mean(daily_income)) 

mod_data |>
  ggplot(mapping=aes(y=avg_mortality,x=avg_income)) +
  geom_point(color='steelblue') +
  theme_classic() +
  labs(title = "Child Mortality vs. Daily Income",
       x = "Daily Income (2017 USD)",
       subtitle = "Child Mortality (per 1000 born)",
       y="")

mod <- lm(avg_mortality~log(avg_income), data=mod_data)
```
$$mortality=58.486 + -1.037  income + \epsilon$$
Interpretation of Coefficients: 

 * $\beta_0$: when daily average income equals 0 in a country, on average child mortality rate per 1000 babies born will be 58.486. 
 * $\beta_0$: If daily average income in a country increases by 1 2017 U.S. dollar, then child mortality rate will decrease by 1.037 on average.
 
   From the linear regression model, we first observe the negative relationship between average income and child mortality, in which higher income levels are associated with lower child mortality rates. The p-values for the slope and intercept are less than 0.001, indicating statistical significance for both coefficients. Also , the R squared value of 0.3229 indicates that approximately 32.29% of the variability in child mortality can be explained by the average income, suggesting that there are certainly other factors that contribute to variability in child mortality that is not captured in this model.
 
# Section 2.3: Model Fit

```{r}

response_variance <- var(mod_data$avg_mortality)
fitted_values <- fitted(mod)
fitted_variance <- var(fitted_values)
residuals <- resid(mod)
residual_variance <- var(residuals)

variances <- data.frame(
  Measure = c("Response Variance", "Fitted Variance", "Residual Variance"),
  Variance = c(response_variance, fitted_variance, residual_variance)
)

kable(variances, format = "html") |>
  kable_styling()


```
  The r-squared value represents the proportion of the variability in the response variable that is explained by the explanatory variable. In our model, the r-squared value is 0.3229, which means that approximately 32.29% of the variability in child mortality rates is explained by the average daily income of the country for that year. This indicates that our model explains a statistically significant portion of the variability, supporting our hypothesis that there is an inverse relationship between child mortality rates and average daily income albeit moderately weak. However, it also indicates that there is still a substantial amount of variability (67.71%) that is not explained by the model, suggesting that other factors also play an important role in determining child mortality rates that are not represented in the relationship between these two variables.
  
  
# Section 3.1: 


```{r}
pred_mod <- predict(mod)
est_sigma <- summary(mod)$sigma
```

```{r}
rand_error <- function(x, mean = 0, sd){
  
  errors <- rnorm(length(x), mean = mean, sd = sd)
  x_with_error <- x + errors
  
  return(x_with_error)
  
}
```

```{r}
sim_response <- tibble(sim_mod = rand_error(pred_mod,
                                            sd = est_sigma),)
```

```{r}
full_data <- mod_data |> 
  select(avg_mortality, avg_income) |> 
  bind_cols(sim_response)
```

Using our linear regression model, we generated predictions for child mortality and added random error to them. below is the side by side plots of our simulated response on the left and our acutal data on the right.

```{r}
plot1 <- full_data |>
  ggplot(mapping=aes(y=sim_mod,x=avg_income)) +
  geom_point(color='steelblue') +
  theme_classic() +
  labs(title = "Simulated Response vs. Daily Income",
       x = "Daily Income (2017 USD)",
       subtitle = "Child Mortality (per 1000 born)",
       y="")
plot2 <- mod_data |>
  ggplot(mapping=aes(y=avg_mortality,x=avg_income)) +
  geom_point(color='steelblue') +
  theme_classic() +
  labs(title = "Child Mortality vs. Daily Income",
       x = "Daily Income (2017 USD)",
       subtitle  = "Child Mortality (per 1000 born)",
       y = ""
       )
grid.arrange(plot1, plot2, ncol = 2)
```
Our Simulated data and our actual observations follow a similar downward slopping pattern, with our observations being more concentrated around low daily income and high child mortality. One of the main differences between our simulated and observed is that our actual observations have some outliers with both a very high average daily income and high child mortality rate. Overall, our simulated data ligns up very well with our observed data. But, to see if our model holds up, we will generate one thousand simulated datasets to perform predictive checks.

# Section 3.2: Generating Multiple Predictive Checks 

```{r}
# Function to run a regression and extract R-squared value
run_simulation <- function(pred_mod, est_sigma, mod_data) {
  # Generate simulated response
  sim_mod <- rand_error(pred_mod, sd = est_sigma)
  
  # Remove NAs and bind with observed data
  full_data <- mod_data %>% 
    select(avg_mortality, avg_income) %>% 
    bind_cols(tibble(sim_mod = sim_mod)) %>%
    na.omit()
  
  # Fit the regression model
  sim_lm <- lm(avg_mortality ~ sim_mod, data = full_data)
  
  # Return the R-squared value
  summary(sim_lm)$r.squared
}
```

```{r}
r_squared_values <- map_dbl(1:1000, ~ run_simulation(pred_mod, est_sigma, mod_data))
r_squared_df <- tibble(r_squared = r_squared_values)
```

We generated 1000 simulated data sets in order to see if they could have been observed based on our regression model. We then regressed our observed dataset against each simulated response, then took the $R^2$ value from each regression and plotted them below:

```{r}
ggplot(r_squared_df, aes(x = r_squared)) +
  geom_histogram(binwidth = 0.01, fill = 'steelblue', color = 'black') +
  theme_classic() +
  labs(title = "Distribution of R^2 Values from Simulated Regressions",
       x = "R^2 Value",
       y = "Frequency")

```
Our simulated $R^2$ values follow a normal distribution, which should be expected given that we added normally distributed random error. It is centered around a $R^2$ of .2, meaning that on average our simulated data sets derived from our regression explains 20% of the variability in our observed data. when creating simulated data sets, our regression leaves 80% of the variability in our original data unexplained.

# Conlusion

The hypothetized relationship explored in this report was the relationship between average daily income and child mortality across 185 countries. We suspected that as average daily income increased, child mortality would decrease. In order to test this relationship, we created a linear regression model, and found a negative relationship between the two variables.
